Do moral communities have a spatial dimension? A spatial exploratory analysis of places of worship and violent crime in the city of Recife, Brazil

### 2019-10-15 Moral_Communities_and_Crime_Title_Page_v1.tex
First complete draft of paper and title page

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

\author{  Aye B. Cee \and  Dee E. Eff \and  Gee H. Eye \and  Jay K. Elle \and  }

\institute{
        Aye B. Cee \at
     A Department, University of XXX \\
     \email{\href{mailto:abc@example.edu}{\nolinkurl{abc@example.edu}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Dee E. Eff \at
     A Department, University of XXX \\
     \email{\href{mailto:def@example.edu}{\nolinkurl{def@example.edu}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Gee H. Eye \at
     A Department, University of XXX \\
     \email{\href{mailto:ghi@example.edu}{\nolinkurl{ghi@example.edu}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Jay K. Elle \at
     Some School, University of WWW \\
     \email{\href{mailto:jkl@school.edu}{\nolinkurl{jkl@school.edu}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@

\hypertarget{intro}{%
\section{Introduction}\label{intro}}

Violent crime is a widespread phenomenon with negative impacts on many
spheres of life and society. Although the rate of homicides worldwide
has grown at a slower rate than the population in the past few decades,
the number of people killed in homicides still increased from 362,000 in
1990 to 464,000 in 2017 (United Nations Office on Drugs and Crime
2019a). Furthermore, variations in the prevalence of violent crime tend
to be extremely uneven. The Americas, with a population of approximately
793.8 million people in 2019 (or about 10.3\% of the world population),
accounted for approximately 173,000 homicides in 2017, or approximately
37.3\% of all homicides in the world. Within the Americas, Brazil (along
with Venezuela, Colombia, and Mexico) is one of the largest countries in
the region with high homicide rates. And there are notable variations at
the sub-national and intra-metropolitan levels as well. In fact,
understanding the variations in the prevalence of violent crime is
recognized as key to achieving policy goals:

\begin{quote}
\begin{quote}
High levels of homicidal violence are concentrated in geographic and
demographic ``pockets'', so achieving target 16.1 of the Sustainable
Development Goals requires interventions within the specific regions,
countries, communities and population groups that are most at risk
(United Nations Office on Drugs and Crime 2019a, 35)
\end{quote}
\end{quote}

This concern is not iddle: the endemic malady of violent crime
represents an important factor that threatens to derail sustainable
development goals. Violent crime, furthermore, tends to be more accute
in those regions most in need of development. On the one hand, violent
crime represents an economic and social drag (Becker and Kassouf 2017).
The impacts are multifaceted, since they affect people's well-being,
either through the loss of human life, mental health, and limitations in
the right to public spaces (Doran and Burgess 2011), or through
disturbances in schooling and academic achievement (United Nations
Office on Drugs and Crime 2019b). In turn, these negative effects
combine to create the unfortunate conditions that tend to breed criminal
behavior, thus creating a vicious cycle of economic disadvantage and
violence (United Nations Office on Drugs and Crime 2019b). It is not
surprising, given these pecularities, that researchers have joined calls
for increased attention to the study of the patterns of violent crime in
Low and Middle Income Countries (LMIC; Murray, Cerqueira, and Kahn
2013).

Added to the scenario, formal social control institutions in countries
like Brazil leave something (or much) to be desired: while in practice
they should inhibit criminal behavior, they suffer from deep
deficiencies that end up reducing the deterrent power of the justice
system. Serious institutional problems include the inefficiency of the
police, the lack of national legislation, the glacial pace of judicial
processes, and the weak situation of the prison system in the country
(Menezes et al. 2013). This makes it even more urgent to understand the
role of \emph{informal social controls}, i.e., the ability of community
organizations such as schools, clubs, and neighborbood associations to
suppress crime by strengthening the capabilities of neighbors to control
inappropriate behavior (Groff 2015).

In order to understand the factors that can potentially help to deter
crime, beyond formal control, it is important to identify empirical
regularities. Criminological factors include concrete elements (such as
te presence of arms or drugs, or elements of the built environment), and
also figurative factors, which include the soci costs of deviant
behavior as well as family supervision. Accordingly, a number of studies
have investigated various aspects of the environment and neighborhood
design (e.g., Foster, Giles-Corti, and Knuiman 2010; He, Páez, and Liu
2017; Loukaitou-Sideris et al. 2001), whereas other studies have focused
on exposure to environmental attributes that signal weakened norms, such
as liquor and tobacco outlets (e.g., Brower and Carroll 2007; Deryol et
al. 2016; Lipton et al. 2008; Quick, Law, and Luan 2017).

Yet another fruitful avenue for research, and one that has only recently
begun to be explored, is the presence of environmental attributes that
can help to reinforce moral norms, such as schools and churches (e.g.,
Abdullah et al. 2018; Davignon and Thomson 2015; Furr-Holden et al.
2010; Traunmuller 2011). It is thus that in a recent paper, Warner and
Konkel (2019) note that the role of places of worship, as distinct
entities from the members of the congregations, have received less
attention in empirical and theoretical research for their deterrence
potential. The role of these institutions might be particularly
important in places where formal state institutions lack the means or
the will to enforce norms in a consistent way - as is the case in Brazil
(Garmany 2014).

With the above considerations in mind, the objective of this paper is to
investigate whether and how places of workship correlate spatially with
criminal events, or in other words, to investigate whether their signals
as moral communities have a discernible spatial dimension. The case
study is the city of Recife in the state of Pernambuco, in Brazil's
Northeast. Recife is a large and important metropolitan area in a
historically poor region, and afflicted by high levels of violent crime,
having the dubious honor of being one of five state capitals with the
highest rates of homicide in the period under study (Menezes et al.
2013). The empirical strategy is to use spatial analysis to explore the
potential geographical relationships between violent criminal events, on
the one hand, and places of worship and a selection of commercial
establishments that serve as controls, on the other. Dissagregated data
allows us to analyze the relevant point patterns at a very high level of
resolution.

After this introductory section, the rest of this paper is structured as
follows: Section \ref{background} reviews some relevant theoretical
perspectives on religion and crime; then Section \ref{methods} presents
the empirical strategy used in the work. Section \ref{case} presents the
context as well as the data used in the study; Section \ref{results}
presents the analysis, whereas the results are discussed in Section
\ref{discussion}. Finally, some concluding remarks are offered in
Section \ref{conclusions}.

\hypertarget{background}{%
\section{Background: Theoretical Perspectives on Religion and
Crime}\label{background}}

Why do humans behave morally? For millennia, it has been the role of
religion to provide the basic tenets of morally acceptable behavior:
thou shalt not kill et al., and their equivalent in many world religions
(e.g., Donovan 1986). Enforcement of such tenets implies different
mechanisms, including \emph{sin}, \emph{haram}, \emph{karma}, and
\emph{tapu}, with punishment delivered by hellfire and exile to
\emph{Jahannam} or \emph{Gehenna}, to mention just some choice places of
torment. In addition to acting as \emph{de facto} social policy for much
of history, many of these religious tenets still have the force of law
in many cultures and regions - and even where they do not, they are held
by some researchers and policy experts to be helpful complements to
reduce crime in any case (e.g., Durrant and Poppelwell 2017; Johnson
2011). The hypothesis that religion can act as a factor that deters and
reduces criminal behavior, therefore, has prompted the scientific study
of the effectiveness of religion on moral behavior (Hoffmann 2015).

An early effort to theorize the effect of religion on criminal behavior
was the \emph{hellfire hypothesis} of Hirschi and Stark (1969). The
focus of this hypothesis is the threat of extratemporal (and possibly
eternal) punishment, and how this threat can deter believers from
commandment-breaking (also see Pascal's wager). Accordingly, Hirsch and
Stark (1969) posit that negative correlations between crime and religion
are a consequence of a sense of commitment with normative values, a
commitment that is ritually reinforced in a regular fashion (e.g., the
liturgical rite of peace in Catholic Mass). Rohrbaugh and Jessor (1975),
for instance, note that by attributing to divinity (or some other
supra-terrenal entity) the supreme force of punishment, religion helps
to build a sense of obduracy, or ``hardening'' against temptation.

Despite its intuitive appeal, research has not provided much direct
evidence for the hypothesis of hellfire (Hoffmann 2015, 1). In a bizarre
twist, even mainstream doctrines (such as absolution) may in fact
encourage beliefs that neutralize the fear of \emph{terrenal}
punishment, and thus turn out to be criminogenic (e.g., Topalli,
Brezina, and Bernhardt 2013). More deviant cases can even recast
criminal behavior as a form of spiritual insurgency (Chesnut 2017).
Counterexamples like this notwithstanding, researchers remain open to
the possibility that ``religion may still serve as a social control
mechanism by encouraging conventional beliefs, monitoring behaviors,
enhancing family attachments, or providing conventional activities''
(Hoffmann 2015, 1). In this way, the hypothesis of \emph{moral
communities} (Stark 1996) recognizes that social integration is
essential for increasing social control, thus reducing the practice of
behavior that is not in accordance with current norms. Rohrbaugh and
Jessor (1975) also emphasize that religion acts as social control since
it defines what is an appropriate attitude according to moral values,
thus making moral communities a close relative of the concept of
\emph{informal social control}, defined as ``the ability of social
groups or institutions to make norms or rules effective'' (Reiss 1951,
196; cited in Groff 2015, 91).

The hypothesis of moral communities has over the years been used to
examine a variety of outcomes of interest. Recent examples include
Stroope and Baker's (2018) exploration of religiosity and self-rated
health; Davignon and Thomson (2015) with their research on institutional
context and the religiosity of students; and the study of religion as a
source of trust of Traunmuller (2011). This is in addition, of course,
to numerous studies on criminal behavior such as Eitle (2011), an author
who explored the deterrence power of religion on gambling; Lee and
Bartkowski's (2004) investigation of juvenile homicide in rural areas;
and the research of Regnerus (2003) on adolescent delinquency. The
hypothesis of moral communities as a form of informal social control, on
the other hand, has been less studied from a geographical perspective,
and it is only recently that has attracted the attention of researchers.
Groff (2015), for instance, discusses informal social control as a
phenomenon that can plaussibly operate at different geographical scales,
from the level of the home and family, through the street block and
neighborhood, possibly to other scales such as the county. In this way,
Nie and Yang (2019) remark on a recent paper on the lack of research
conducted to study how the religious context of a geographical area
(e.g., a county) may influence (youth smoking) behavior (p.~2). This
point is echoed by Warner and Konkel (2019), who moreover recommend the
use of even smaller units of analysis (e.g., Census Block Groups), in an
effort to reduce aggregation bias (see Hipp 2007), but more importantly
because the processes defined by social disorganization theory are
thought to occur at relatively small geographical scales.

The preceding review, as well as other research (2019), makes clear the
need for research at various geographical scales. From a theoretical
perspective, the social mechanisms that underpin the hypothesis of moral
communities and informal social control can happen at various
geographical scales, some of which previous research has already
addressed: the work of Nie and Yang (2019), for one, highlights the role
of processes detectable at a relatively high level of aggregation,
whereas Warner and Konkel (2019), looking specifically at the effect of
places of worship, bring the study closer to the level of the
neighborhood. In what follows, we propose to adopt a much more
disaggregated approach by considering the intensity of criminal events
with respect to places of worship. This requires micro-level data both
of crimes and places of worship, and the use of appropriate spatial
analytical tools for point patterns, as discussed next.

\hypertarget{methods}{%
\section{Empirical Strategy}\label{methods}}

As noted above, previous research that has investigated the presence of
churches from a geographical perspective has used data aggregated at
different scales. In the case of Nie and Yang (2019), the unit of
analysis was the county, whereas Warner and Konkel (2019) used Census
Block Groups, a much smaller unit of analysis. Research in spatial
criminology, on the other hand, has studied criminal events as point
patterns since at least the work of Levine et al.~(1986) investigated
the concentration of criminal events in the proximity of bus stops.
Since then, many other works have applied tools of spatial point pattern
analysis to investigate the empirical properties of the distribution of
crime. This includes Craglia et al.~(2000) who used the Ripley's
K-function (Ripley 1976) to investigate clustering processes, and
Rogerson and Sun (2001) who applied nearest neighbor techniques to the
study of arson. More recent studies of crime as point patterns are
Nakaya and Yano (2010), Kiani et al.~(2015), and Malleson and Andresen
(2015).

Readers interested in point pattern analysis techniques are urged to
consult the still valuable Bailey and Gatrell (1995) or for an
up-to-date and in-depth coverage of the topic, the excellent text by
Baddeley et al.~(2015). In this paper we will concentrate on one
property of point processes, namely the intensity. Suppose that the
outcome of interest is the number of events per unit area
\(Y(\mathcal{A})\). This could be, for instance, the number of criminal
events observed in an arbitrary area. The intensity of a point pattern
then is as follows: \[
\lambda(s)=\lim_{\text{d}s \to 0}\frac{E[Y(\text{d}s)]}{\text{d}s}
\]

where \(\lambda(s)\) is the intensity of the process at point \(s\),
given by the expected number of events in a small area \(\text{d}s\)
around \(s\), as the area becomes arbitrarily small. If the process is
homogeneous (i.e., the intensity is a constant over space), an apt
estimator of the intensity is the global intensity, simply the number of
events divided by the area of the region under analysis: \[
\hat{\lambda} = \frac{n}{|\mathcal{A}|}
\]

When the process is not homogeneous, other estimators are more
appropriate. In the analysis to follow we use two estimators of
intensity: conditional quadrat counts and relative distribution
estimate. These techniques are briefly discussed next.

\hypertarget{conditional-quadrat-counts}{%
\subsection{Conditional Quadrat
Counts}\label{conditional-quadrat-counts}}

Quadrat counts is a relatively simple technique to analyze spatial
variations in the intensity of a point process. It operates by
partitioning the region under study \(\mathcal{A}\) into subregions
\(\mathcal{A}_1,\cdots,\mathcal{A}_m\). These subregions are mutually
exclusive, and their union is identical to \(\mathcal{A}\). In the
simplest case, the subregions have equal area. The number of events
within each subregion (i.e., \(n_{\mathcal{A}}\)) divided by the area of
the subregion (i.e., \(|\mathcal{A}|\)) is a local estimate of the
intensity: \[
\lambda(\mathcal{A}_i)=\frac{n_{\mathcal{A}_i}}{|\mathcal{A}_i|}
\]

A test for homogeneity consists of assessing whether the intensity of
the point process at the quadrats is uniform: \[
\lambda(\mathcal{A}_1)=\cdots=\lambda(\mathcal{A}_m)
\]

Using quadrats of equal size is a convenient simplication, but in
principle the areas could be different - in which case the count of
events would be proportional to the area of the quadrat under
homogeneity. An interesting variation of this technique is conditional
quadrat counts, whereby the partition of regions
\(\mathcal{A}_1,\cdots,\mathcal{A}_m\) is done to reflect an underlying
covariate of interest, say \(Z\). By introducing a covariate as a
partitioning criterion, it becomes possible to calculate estimators of
the intensity of quadrats at different levels of the value of the
covariate. For exploratory purposes, we can plot average intensity by
quadrat, and compare to the global intensity.

\hypertarget{relative-distribution-estimate}{%
\subsection{Relative Distribution
Estimate}\label{relative-distribution-estimate}}

Conditional quadrat counts allow us to explore whether the intensity of
the process depends on a covariate \(Z\). A different way of expressing
this is as follows: \[
\lambda(s)=\rho\big(Z(s)\big)
\]

In this case, the intensity is a function \(\rho\) that maps how the
intensity depends on covariate \(Z\). Non-parametric estimation of
\(\rho\) uses the ratio of the density of covariate values at the
locations of the point process, relative to the \emph{spatial
distribution function} \(G\), the density of covariate values at random
locations (see Baddeley, Rubak, and Turner 2015, 179).

The density of covariate values at the locations of the is obtained by
means of a kernel density estimator, for example: \[
\lambda_0(s) = \frac{1}{|\mathcal{A}|}\sum_{i = 1}^n\kappa(Z(s_i) - z)
\] where \(s_i\) are the locations of the points.

On the other hand, \(G\) which is the cumulative distribution function
of \(Z\) at random point \(S\) uniformly distributed in \(\mathcal{A}\)
: \[
G(z) = \frac{1}{|\mathcal{A}|} \int_{\mathcal{A}}\text{I}\{Z(s)\leq z\}\text{d}s
\] where \(\text{I}\) is an indicator function that takes the value of
\(1\) if the argument is true, and the value of \(0\) otherwise. In
practice, the spatial distribution function is approximated based on a
discretization of space using a fine grid of pixels as follows: \[
G(z) = \frac{\#\{\text{pixels }s:Z(s)\leq z\}}{\# \text{pixels}}
\]

Therefore, an estimator of \(\rho\) is as follows: \[
\rho(z) = \frac{1}{|\mathcal{A}|G'(z)}\sum_{i=1}^n\kappa(Z(s_i) - z)
\] where the derivative \(G'\) can be approximated by differentiating a
smoothed estimated of \(G\). Other estimators are discussed by Baddeley
et al.~(Baddeley, Rubak, and Turner 2015, 180).

It is possible to adjust the relative distribution estimate by means of
a baseline; a baseline in this case can be a function of other
covariates that might confound the estimates of the relative
distribution, so that the relative intensity \(\lambda(s)/B(s)\) can be
assumed to depend only on covariate \(Z\). Therefore: \[
\lambda(s) = \rho(Z(s))B(s)
\] where \(B(s)\) is a baseline function. It can be seen that
\(\rho(z)=1\) is the baseline intensity; values of \(\rho(z)>1\)
correspond to intensities higher than the baseline \emph{as a function
of} \(Z(u)\), whereas values of \(\rho(z)<1\) correspond to intensities
lower than the baseline, again as a function of \(Z(u)\).

\hypertarget{case}{%
\section{Case Study: Context and Data}\label{case}}

The study is of the city of Recife, the capital of the state of
Pernambuco in the Northeast region of Brazil. With a population of
1,550,390 million inhabitants in 2010, Recife is one of the main
Brazilian metropolises, exerting a great economic influence in
neighboring regions. However, the city experiences a serious problem
with violent crime, and has the dubious honor of being one of the five
capital cities in Brazil with the highest homicide rates in the period
under study (Menezes et al. 2013).

In the current context, the term ``violent crime'' is an umbrella for
several forms of infractions to the penal code. Following
recommendations of the National Secretariat of Public Security of the
Ministry of Justice of 2006, these are Violent Lethal and Intentional
Crimes (VLIC; which includes intentional homicide), theft followed by
death (robbery), and corporal injury followed by death. The data on LIVC
were extracted from the Police Information System of the Secretariat of
Social Defense of Pernambuco (INFOPOL / SDS-PE), which is the most
reliable, detailed, and comprehensive information on violent deaths in
the region.

\begin{table}

\caption{\label{tab:table-descriptive-statistics}\label{tab:descriptive-statistics}Characteristics of Violent Crime in Recife, July, 2008 - June, 2009}
\centering
\begin{tabular}[t]{lr}
\toprule
 & Percentage\\
\midrule
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Gender of Victim}}\\
\hspace{1em}Man & 91.35\\
\hspace{1em}Woman & 8.65\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Type of Crime}}\\
\hspace{1em}Murder & 97.82\\
\hspace{1em}Robbery & 2.00\\
\hspace{1em}Body injury followed by death & 0.18\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Ethnicity of Victim}}\\
\hspace{1em}Black and White & 95.11\\
\hspace{1em}Yellow and White & 1.27\\
\hspace{1em}Not Reported & 3.62\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Age of Victim}}\\
\hspace{1em}1 to 12 years & 0.18\\
\hspace{1em}13 to 17 years & 11.31\\
\hspace{1em}18 to 30 years & 61.70\\
\hspace{1em}31 to 65 years & 25.84\\
\hspace{1em}65 years and older & 0.73\\
\hspace{1em}Not Reported & 0.24\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Weapon Used}}\\
\hspace{1em}Firearm & 87.51\\
\hspace{1em}Other & 12.49\\
\bottomrule
\end{tabular}
\end{table}

The data are organized at the individual level and it is possible to
obtain information about the location, day of the week, day of the
month, period of the day, as well as gender, age, and race of the
victim. The database used in this study comprises the period from July
1, 2008 to June 30, 2010. Some descriptive statistics regarding this
dataset are reported in Table 1, where it can be seen that overall,
about 91\% of the victims of LIVC in the period analyzed were men, while
approximately 98\% of violent crimes were homicides. In addition, most
of the victims were black or brown, and the youth population between the
ages of 18 and 30 is the most affected by violent crime. Lastly, it
should be noted that about 88\% of the criminal events under analysis
involved firearms. Figure \ref{fig:plot-crime} shows the spatial
distribution of the 1,657 LIVC crimes that occurred in the city of
Recife between July, 2008 and June, 2010.

\begin{figure}

\includegraphics{Moral_Communities_and_Crime_v1_files/figure-latex/fig-plot-crime-1} \hfill{}

\caption{\label{fig:plot-crime}Location of Lethal and Intentional Violent Crime in Recife, July, 2008 - June, 2009}\label{fig:fig-plot-crime}
\end{figure}

Information about places of interest was obtained from the National
Register of Addresses for Statistical Purposes (CNEFE - Census 2010),
which lists 78,056,411 urban and rural addresses, distributed among the
316,574 census tracts. This is the first database of its kind produced
by IBGE, and the first version was produced at the time of the 2000
Census. The way addresses are described in the National Register is very
rich, and it is possible to identify the names of the places of worship
including their denomination. Georeferencing was used to geolocate each
place of worship. In this way, a total of 1,719 places of worship were
geolocated in the city of Recife.

In addition to places of worship, the National Register of Addresses for
Statistical Purposes was queried to extract facilities other than places
of worship. As discussed above, the idea is to identify points of
reference that can be used as controls, having a neutral morality
profile. For the sake of the present study, we selected pharmacies, ice
cream shops, and bakeries to construct our control group. These three
types of establishments comply with the criteria of being morally
neutral and having a spatial distribution commensurate with places of
worship. Figure 2 shows the spatial distribution of the control
establishments, namely pharmacies, bakeries, and ice cream shops in the
city of Recife. Note the similarity between the maps. As expected, there
are differences in the number of points, but the locations of cases and
controls are quite similar.

\begin{figure}

\includegraphics{Moral_Communities_and_Crime_v1_files/figure-latex/fig-plot-events-1} \hfill{}

\caption{\label{fig:plot-events}Location of Places of Worship and Commercial Establishments in Recife}\label{fig:fig-plot-events}
\end{figure}

\hypertarget{results}{%
\section{Analysis and results}\label{results}}

The analysis is implemented using the \texttt{R} statistical computing
language, and documents to replicate the analysis, in addition to all
data necessary, can be obtained from this anonymous shared folder:

\url{https://drive.google.com/open?id=1tuJM4Mhi0Ftq3ZEwjv6RP9vO3veGjIR_}

\hypertarget{socio-economic-deprivation}{%
\subsection{Socio-Economic
Deprivation}\label{socio-economic-deprivation}}

The first step in our analysis is to obtain an indicator of
Socio-Economic Deprivation (SED). Socio-economic deprivation is known to
correlated positively with crime, as it often reflects relevant
criminogenic factors such as poverty and family disruption (He et al.
2015). In the present case, we have for the city of Recife information
about the variables shown in Table \ref{tab:sed-descriptive-statistics}
at the level of \emph{setores}, a small Brazilian census geography.

As seen in the table, Recife is a city with large socio-economic and
demographic disparities; for example, the \emph{setor} with the highest
median income has a median income that is 6,517\% higher than the median
income in the \emph{setor} with the lowest median income. Whereas there
are \emph{setores} where the proportion of population who are unemployed
is zero, there are \emph{setores} where almost 60\% of the population
are unemployed. Likewise, there are \emph{setores} where almost 70\% of
the population lives in poverty. In addition to these economic
indicators, two variables are used to represent family disruption, the
proportion of families whose head is a single mother, and the proportion
of families whose head is a \emph{young} single mother, that is, a woman
between the ages of 15 and 25. As can be seen, there are \emph{setores}
with approximately 70\% of households led by single mothers, and of
these, over 21\% are led by younger women, indicating a high degree of
family disruption.

For the analysis, we use Principal Component Analysis, a data reduction
technique, to obtain an indicator of Socio-Economic Deprivation,
essentially the first principal component in the output. The loadings of
this factor (Table \ref{tab:sed-descriptive-statistics}) indicate that
high Socio-Economic Deprivation is a combination of (in terms of
importance): high levels of poverty, high levels of unemployment, and
low median income, followed by high levels of family disruption, in
particular proportion of families led by young single mothers. This
factor accounts for almost 54\% of the variance..

\begin{table}

\caption{\label{tab:table-sed-descriptive-statistics}\label{tab:sed-descriptive-statistics}Descriptive statistics of some key socio-economic and demographic variables in Recife}
\centering
\begin{tabular}[t]{lccccccc}
\toprule
Variable & Min & 2nd Quartile & Median & Mean & 3rd Quartile & Max & \makecell[l]{PC Factor 1\\ Loadings}\\
\midrule
Median Income & 122.05 & 337.40 & 581.67 & 1031.09 & 1283.79 & 7954.19 & -0.42\\
Proportion Unemployed & 0.00 & 0.05 & 0.10 & 0.10 & 0.14 & 0.59 & 0.56\\
Proportion Poverty & 0.00 & 0.05 & 0.11 & 0.12 & 0.17 & 0.63 & 0.58\\
Proportion Single Mother & 0.00 & 0.33 & 0.37 & 0.37 & 0.41 & 0.70 & 0.17\\
Proportion Young Single Mother & 0.00 & 0.01 & 0.02 & 0.02 & 0.02 & 0.21 & 0.38\\
\bottomrule
\multicolumn{8}{l}{\textit{Note: }}\\
\multicolumn{8}{l}{The variable for young single mothers is for women aged 15-25}\\
\multicolumn{8}{l}{The first principal component accounts for 53.91\% of the variance}\\
\end{tabular}
\end{table}

Figure \ref{fig:plot-sed-as-quintiles} displays the geography of
Socio-Economic Deprivation in Recife, after classifying \emph{setores}
by quintiles, whereby the ``High'' class corresponds to \emph{setores}
in the top 20\% of the Socio-Economic Deprivation indicator, and the
``Low'' class corresponds to \emph{setores} in the bottom 20\% of the
Socio-Economic Deprivation indicator. The figure shows a veritable
mosaic of affluence and deprivation, with high deprivation areas
directly in contact, and in some cases almost completely surrounded, by
low deprivation areas. This geographical pattern of unequality, on the
other hand, seems to be characteristic of Brazilian metropolitan
regions, where enclaves of wealth and \emph{favelas} (i.e., urban slums)
can be found in close proximity (see for example Feitosa et al. 2007).

\begin{figure}
\centering
\includegraphics{Moral_Communities_and_Crime_v1_files/figure-latex/plot-sed-as-quintiles-1.pdf}
\caption{\label{fig:plot-sed-as-quintiles}Socio-Economic Deprivation in
Recife classified by quintiles}
\end{figure}

\hypertarget{conditional-quadrat-analysis}{%
\subsection{Conditional Quadrat
Analysis}\label{conditional-quadrat-analysis}}

Once the indicator of Socio-Economic Deprivation has been obtained, as
outlined above, it can be used for conditional quadrat analysis. As
previously described, the quadrats are irregular areas that are defined
based on a covariate, in this case the quintiles of the Socio-Economic
Deprivation indicator. In essence, this entails counting the number of
events at each level of the covariate, and then calculating an estimator
of the intensity.

Figure \ref{fig:plot-crime-quadrat} shows the results of conditional
quadrat counts for Lethal and Intentional Violent Crime. The dotted line
indicates the global intensity, simply the total number of criminal
events divided by the area of the region. It can be seen that the
intensity of crime clearly increases with Socio-Economic Deprivation,
and the intensity in areas with the highest levels of Socio-Economic
Deprivation is 208.7065292 \% higher than in places with the lowest
levels of Socio-Economic Deprivation. This is as expected, as research
has consistently found correlations between Socio-Economic Deprivation
and a variety of negative health outcomes, including homicide (Ichihara
et al. 2018). A \(\chi^2\) test of independence on the quadrat counts
yields a value of \(p\leq 0.0001\) (four degrees of freedom), which
comfortably rejects the null hypothesis of homogeneity of intensity of
crime by level of Socio-Economic Deprivation.

\begin{figure}
\centering
\includegraphics{Moral_Communities_and_Crime_v1_files/figure-latex/plot-crime-quadrat-1.pdf}
\caption{\label{fig:plot-crime-quadrat}Intensity of crime by level of
Socio-Economic Deprivation; the dotted line indicates the global
intensity of crime}
\end{figure}

Next, we repeat the analysis of conditional quadrat counts, but now for
the six types of events of reference. As discussed above, this included
three types of places of worship which are of interest from the
perspective of moral communities, and three types of commercial
establishments that we posit are morally neutral. The results of
estimating the intensity of the process by means of conditional quadrat
counts are shown in Figure \ref{fig:plot-events-quadrat}.

The first thing that we note are the variations in global intensity of
places of worship, with Evangelical places of worship being the most
intense of the three. There are no clear variations in the locational
pattern of Catholic places of worship by level of Socio-Economic
Deprivation. Evangelical places of worship, on the other hand, tend to
be found more frequently in places with middle-low and middle levels of
Socio-Economic Deprivation, a pattern also discernible, albeit barely,
for Spiritist churches. Interestingly (see Table
\ref{tab:quadrat-test-statistics}) \(\chi^2\) tests of independence on
the conditional quadrat counts fail to reject the hypothesis of
homogeneity for Catholic and Spiritist churches, and it is only in the
case of Evangelical places of worship that we detect the possibility of
locational patterns that vary by level of Socio-Economic Deprivation.

Three types of commercial establishments also appear to display
inhomogeneous locational patterns (Table
\ref{tab:quadrat-test-statistics}), with the \(\chi^2\) test of
independence emphaticaly rejecting the null hypothesis. As seen in
Figure \ref{fig:plot-events-quadrat}, ice cream shops tend to be found
more frequently in places with middle-low Socio-Economic Deprivation,
and less frequently in places with middle-high Socio-Economic
Deprivation. Pharmacies show a clearer trend, with locational patterns
that tend to favor places with low Socio-Economic Deprivation. Finally,
supermarkets are found less frequently in places with middle-high and
high Socio-Economic Deprivation.

\begin{figure}
\centering
\includegraphics{Moral_Communities_and_Crime_v1_files/figure-latex/plot-events-quadrat-1.pdf}
\caption{\label{fig:plot-events-quadrat}Intensity of places of worship
and commercial establishments by level of Socio-Economic Deprivation;
the dotted lines indicates the respective global intensities}
\end{figure}

\begin{table}

\caption{\label{tab:table-quadrat-test-statistics}\label{tab:quadrat-test-statistics}$\chi^2$ tests of independence of conditional quadrat counts}
\centering
\begin{tabular}[t]{lccc}
\toprule
Event & Statistic & Degrees of Freedom & p-value\\
\midrule
Catholic & 6.913 & 4 & 0.2811\\
Evangelical & 38.669 & 4 & $<0.001$\\
Spiritist & 9.108 & 4 & 0.1169\\
Ice Crem Shop & 15.108 & 4 & 0.009\\
Pharmacy & 51.869 & 4 & $<0.001$\\
\addlinespace
Supermarket & 23.826 & 4 & $<0.001$\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{relative-distribution}{%
\subsection{Relative Distribution}\label{relative-distribution}}

The preceding analysis provides a valuable backdrop. As seen there,
there is a clear distribution of criminal events, increasing on par with
the level of Socio-Economic deprivation. Of the three different classes
of places of worship, only Evangelical churches display a locational
pattern that is commensurate, with Socio-Economic Deprivation, albeit
the pattern is distinct from criminal events, with Evangelical churches
found less frequently in both high and low Socio-Economic Deprivation
areas.

The analysis does not answer the question, yet, of possible covariations
between crime, on the one hand, and places of worship and commercial
establishments, on the other. In this subsection we implement the
relative distribution estimator of intensity of Lethal and Intentional
Violent Crime. The covariate \(Z(u)\) in this analysis is distance to
one of a place of worship or a commercial establishment: \[
\lambda_k(s)=\rho(Z_k(s))
\] with: \[
Z_k(s) = d_{sk}
\] In the expression above, \(d_{s_k}\) is distance from location \(s\)
to an event of class \(k\), and
\(k=\{\text{Catholic, Evangelical, Spiritist, Ice Cream Shop, Pharmacy, Supermarket}\}\).

The results of this analysis are shown in Figure
\ref{fig:plot-relative-distribution}. The global estimator for the
intensity of crime is shown in this figure as a dotted line. Three types
of places of worship are shown in solid lines, and three types of
commercial establishments are shown in dashed lines. Each relative
distribution function is shown with its corresponding 95\% confidence
bands.

Several interesting things emerge from inspection of this figure. First,
we notice that with the exception of Spiritist churches (which has an
intensity significantly lower than the global intensity at a distance of
zero), the intensity of crime at for other churches and commercial
establishments is close, but higher than, the global intensity of crime.
In general, the intensity tends to increase at short distances; however,
this effect is much more marked for Evangelical places of worship. The
intensity of crime in the proximity of these places of worship grows
rapidly within a short distance, reaching a peak of
\ensuremath{2.4916\times 10^{-5}} at a distance of 64.92 m. At that
distance, the intensity of crime with respect to Evangelical places of
worship is 183.78\% higher than the intensity of Catholic places of
worship; 194.54\% higher than the intensity of crime with respect to
Spiritist churches; 174.19\% higher than the intensity with respect to
ice cream shops; 151.52\% higher than the intensity with respect to
pharmacies; and 177.63\% higher than the intensity with respect to
supermarkets.

>>>>

@@

\author{  Edivaldo Constantino das Neves Júnior \and  Breno Caldas de Araujo \and  Tatiane Almeida de Menezes \and  Antonio Paez \and  }

\institute{
        Edivaldo Constantino das Neves Júnior \at
     A Department, University of XXX \\
     \email{\href{mailto:edivaldoconstantino@gmail.com}{\nolinkurl{edivaldoconstantino@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Breno Caldas de Araujo \at
     A Department, University of XXX \\
     \email{\href{mailto:def@example.edu}{\nolinkurl{def@example.edu}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Tatiane Almeida de Menezes \at
     Programa de Pós-Graduação em Gestão e Economia da Saúde, Universidade
 Federal de Pernambuco \\
     \email{\href{mailto:tatianedemenezes@gmail.com}{\nolinkurl{tatianedemenezes@gmail.com}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
        Antonio Paez \at
     School of Geography and Earth Sciences, McMaster University \\
     \email{\href{mailto:paezha@mcmaster.ca}{\nolinkurl{paezha@mcmaster.ca}}}  %  \\
%             \emph{Present address:} of F. Author  %  if needed
    \and
    }

@@
